Circular Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision of introduction to section 4; minor corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2880 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 5d368970440aac070259cddf23b8bbf386990728

\title{Circular Regression Trees and Forests with an Application to 
    Probabilistic Wind Direction Forecasting}
\Shorttitle{Distributional Trees and Forests for Circular Data}

@@

\Abstract{
As circular data can be found in many different subject areas accounting for periodicity has 
become an important aspect in establishing appropriate statistical models.
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. In order to stabilize and regularize
the model an ensemble of distributional trees is combined yielding a distributional
forest which allows for modeling smooth effects as well.
For illustration, hourly wind direction forecasts are obtained at two airports in Austria
(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
airports and surrounding weather stations.
}
\Keywords{Distributional Trees and Forests; Circular Response; Von Mises Distribution}

@@

A more extensive case study of circular regression trees and forests applied to
probabilistic forecasting of wind direction at the airports of Innsbruck and
Vienna will be presented in this paper. Moreover, their performances are
benchmarked to alternative commonly used approaches.  The theory on
probabilistic circular modeling introducing the von Mises distribution will be
discussed in Section~\ref{sec:prob_circ}. The methodology of circular
distributional trees and forests and their features are introcuded in
Section~\ref{sec:tree_forest}.  After the case study presented in
\ref{sec:wind}, a comprehensive summary and conclusion is given in
Section~\ref{sec:summary}.

@@

\section{Circular regression trees and forests}
\label{sec:tree_forest}
%\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates is to regress the model parameters on covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs. This kind of approach is presented for example
in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. 
%While applying the mixed model 
%requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
%\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
%a circular response variable.
Apart from the methodology developed specifically for circular data more general approaches to 
regressing model parameters on covariates are commonly used such as for example generalized additive 
models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
where each distribution parameter is estimated by its own generalized additive model (GAM). 
%However, an adaption of this methodology to circular data can be very complex due to a
%challenging representation of smooth effects on the unit circle.
%However, an adaption of this methodology to circular data can not be obtained by the straight 
%forward approach of simply wrapping the modeled additive effects
%around the unit circle since the parameters of the von Mises distribution and their 
%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
However, difficulties might arise in case of non-additive changes and specifying a model 
properly can be a very challenging, particularly for a high number of covariates and no information
on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
Accurate forecasts of wind directions are of great importance for risk
management in various fields such as agriculture, industry or energy
production. Thus, for example in order to direct airplanes to a safe landing,
precise knowledge of wind direction at the respective airport for the next hour(s) is highly
desirable. This section provides a comprehensive case study on probabilistic
wind direction forecasting at two airports in Austria with fundamentally
different site characteristics: Innsbruck with a rather unusual location for an
airport, located at the bottom of a deep alpine valley, and Vienna located
within flat terrain in the East of Austria. In addition to the newly
introduced circular distributional trees and forests, the case study employs various
alternative approaches to forecast wind direction. The study is based on
$1$\,h and $3$\,h forecasts employing lagged observations in the vicinity of
the airports.

@@

\subsection{Data}\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min mean
of wind direction measurements at the airports Innsbruck and Vienna on an hourly
temporal resolution. As predictor variables, we use $1$-hourly resolved
$10$\,min mean observations of various meteorological quantities such as wind
direction, wind speed, temperature, air pressure and humidity, all
lagged by one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
the their vicinity. For Innsbruck, measurements of
21~different locations at the airport and along the intersecting valleys are
used, whereas, for Vienna, measurements of 22~locations at the airport and
within an area of about $10$\,km$^2$ are used. A topographical overview of the
airports and their surrounding areas with the station sites employed in this
study is provided in Figure~XXX. In addition, we use derived quantities such
as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. A summary of all predictor variables can be found in
Table~\ref{tab:covariates}.

@@

\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\caption[Table caption text]{Basic covariates together with the type of variations.}
\label{tab:covariates}
\begin{tabular}{l  l}
\hline
Basic covariates & Variations\\
\hline
\emph{dd}: wind direction       & different locations, \\
                                & changes over time(?)\\
\emph{ff}: wind speed           & different locations, \\
                                & changes over time(?), \\
                                & (hourly) maxima\\
\emph{p}: pressure              & different locations, \\
                                & differences between locations, \\
                                & changes over time(?)\\
\emph{t}: temperature           & different locations, \\
                                & changes over time(?)\\
\emph{rf}: relative humidity    & different locations\\
\hline
\end{tabular}
\end{minipage} 
%}
\end{table}

\subsection{Models and evaluation}\label{sec:wind:models}
For a fair evaluation of circular regression trees and forests, and to
investigate whether they can be applied as a reasonable alternative to already
existing approaches, three additional statistical models are employed in this
study for probabilistic forecasting of wind directions. Two of them are based
on existing approaches in the meteorological field while the third is a
state-of-the-art model to forecast circular response variables. The next
paragraphs introduce the reference methods and provide the corresponding model
specifications as employed in this study for all considered models including the
circular tree and forest.

@@

climatological model by using maximimum likelihood estimation and fitting the distribution

@@

\item \emph{Circular regression tree:}
For the circular regression tree introduced in Section~\ref{sec:circtree}, all

@@

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the $5$~validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance. For both stations at both forecast steps,
the circular regression forest overall provides the best predictive
performance, followed by the circular regression tree and, except for the
$3$\,h forecast at Innsbruck, the persistency model. In comparison to the
circular tree and forest, for both stations and forecast steps, the
climatological model and the linear model show clearly higher CRPS values and
hence a lower predictive performance.

@@

The different forecast steps seem to have only a minor effect on the predictive
performance of the climatological model and the linear model at both stations.
However, for the persistency model, at both stations, higher scores for the
$3$\,h forecast (Panels~c,\,d) reveal a lower performance for longer lead
times; this is probably due to minimum 3-hourly instead of 1-hourly lagged
response values employed as covariates in the persistency model. The circular
regression tree and forest seem to partially compensate for the lower skill of
the lagged response values by other covariates, hence their predictive
performance only slightly decreases for the longer lead time. This compensation
is especially evident for Innsbruck, where the performance difference between
the persistency model and the tree-based methods significantly increases from
the 1-hourly to the 3-hourly forecast.

\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  %\hspace{-2cm}
  %\includegraphics{boxplot_crpsraw.pdf}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores for wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    per for each of the $12$~months averaged over the $5$~validation years which yields a total of
    $288$~yearly mean values. The scores are shown for the climatological, 
    persistency and the linear model as well as for the circular regression tree and
    forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}

>>>>

\title{Circular Trees and Forests with an Application to 
    Probabilistic Wind Direction Forecasting}
\Shorttitle{Circular Trees and Forests}

@@

\Abstract{
As circular data can be found in many different subject areas accounting for periodicity has 
become an important aspect in establishing appropriate statistical models.
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a distributional regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. In order to stabilize and regularize
the model an ensemble of distributional trees is combined yielding a distributional
forest which allows for modeling smooth effects as well.
For illustration, hourly wind direction forecasts are obtained at two airports in Austria
(Vienna and Innsbruck) based on a set of meteorological measurements from each of the 
airports and surrounding weather stations.
}
\Keywords{Distributional Trees and Forests; Circular Response; Von Mises Distribution}

@@

A more extensive case study of circular distributional trees and forests applied to
probabilistic forecasting of wind direction at the airports of Innsbruck and
Vienna will be presented in this paper. Moreover, their performances are
benchmarked to alternative commonly used approaches.  The theory on
probabilistic circular modeling introducing the von Mises distribution will be
discussed in Section~\ref{sec:prob_circ}. The methodology of circular
distributional trees and forests and their features are introduced in
Section~\ref{sec:tree_forest}.  After the case study presented in
\ref{sec:wind}, a comprehensive summary and conclusion is given in
Section~\ref{sec:summary}.

@@

\section{Circular distributional trees and forests}
\label{sec:tree_forest}
%\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates is to regress the model parameters on covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs. This kind of approach is presented for example
in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. 
%While applying the mixed model 
%requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
%\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
%a circular response variable.
Apart from the methodology developed specifically for circular data more general approaches to 
regressing model parameters on covariates are commonly used such as for example generalized additive 
models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
where each distribution parameter is estimated by its own generalized additive model (GAM). 
%However, an adaption of this methodology to circular data can be very complex due to a
%challenging representation of smooth effects on the unit circle.
%However, an adaption of this methodology to circular data can not be obtained by the straight 
%forward approach of simply wrapping the modeled additive effects
%around the unit circle since the parameters of the von Mises distribution and their 
%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
However, difficulties might arise in case of non-additive changes and specifying a model 
properly can be a very challenging, particularly for a high number of covariates and no information
on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
Accurate forecasts of wind directions are of great importance for risk
management in various fields such as agriculture, energy production or air safety control. 
For example in order to direct airplanes to a safe landing, precise knowledge of wind 
direction at the respective airport for the next hour(s) is highly desirable and adequate 
prediction methods are required. In this section, a comprehensive case study on probabilistic
wind direction forecasting is provided for two airports in Austria with fundamentally
different site characteristics: Innsbruck with a rather unusual location for an
airport, located at the bottom of a deep alpine valley, and Vienna located
within flat terrain in the East of Austria. In addition to the newly
introduced circular distributional trees and forests, the case study employs various
alternative approaches to forecast wind direction. The study is based on
$1$\,h and $3$\,h forecasts employing lagged observations in the vicinity of
the airports.

@@

\subsection{Data}\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min mean
of wind direction measurements at the airports Innsbruck and Vienna on an hourly
temporal resolution. As predictor variables, we use $1$-hourly resolved
$10$\,min mean observations of various meteorological quantities such as wind
direction, wind speed, temperature, air pressure and humidity, all
lagged by one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
the their vicinity. For Innsbruck, measurements of
21~different locations at the airport and along the intersecting valleys are
used, whereas, for Vienna, measurements of 22~locations at the airport and
within an area of about $10$\,km$^2$ are used. A topographical overview of the
airports and their surrounding areas with the station sites employed in this
study is provided in Figure~XXX. In addition, we use derived quantities such
as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. A summary of all predictor variables can be found in
Table~\ref{tab:data}.

@@

\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\caption[Table caption text]{Overview of the employed data set.}
\label{tab:data}
\begin{tabular}{l  l}
\hline
Data components              & Description \\
\hline
Meteorological variables     & Wind direction, wind speed, air pressure, \\
                             & temperature, relative humidity \\
Derived quantities           & 3-hourly means, minima, maxima; \\
                             & 1-hourly and 3-hourly temporal changes; \\
                             & spatial differences towards the airport\\
Weather stations (Innsbruck) & 21 stations at the airport, \\
                             & Landeck, Kufstein, Steinach, \dots \\
Weather stations (Vienna)    & 22 stations at the airport, \\
                             & \ldots \\
%\emph{dd}: wind direction       & different locations, \\
%                                & changes over time(?)\\
%\emph{ff}: wind speed           & different locations, \\
%                                & changes over time(?), \\
%                                & (hourly) maxima\\
%\emph{p}: pressure              & different locations, \\
%                                & differences between locations, \\
%                                & changes over time(?)\\
%\emph{t}: temperature           & different locations, \\
%                                & changes over time(?)\\
%\emph{rf}: relative humidity    & different locations\\
\hline
\end{tabular}
\end{minipage} 
%}
\end{table}

\subsection{Models and evaluation}\label{sec:wind:models}
For a fair evaluation of circular distributional trees and forests, and to
investigate whether they can be applied as a reasonable alternative to already
existing approaches, three additional statistical models are employed in this
study for probabilistic forecasting of wind directions. Two of them are based
on existing approaches in the meteorological field while the third is a
state-of-the-art model to forecast circular response variables. The next
paragraphs introduce the reference methods and provide the corresponding model
specifications as employed in this study for all considered models including the
circular tree and forest.

@@

climatological model by using maximum likelihood estimation and fitting the distribution

@@

\item \emph{Circular distributional tree:}
For the circular distributional tree introduced in Section~\ref{sec:circtree}, all

@@

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the $5$~validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance. For both stations at both forecast steps,
the circular regression forest overall provides the best predictive
performance, followed by the circular distributional tree and, except for the
$3$\,h forecast at Innsbruck, the persistency model. In comparison to the
circular tree and forest, for both stations and forecast steps, the
climatological model and the linear model show clearly higher CRPS values and
hence a lower predictive performance.

@@

The different forecast steps seem to have only a minor effect on the predictive
performance of the climatological model and the linear model at both stations.
However, for the persistency model, at both stations, higher scores for the
$3$\,h forecast (Panels~c,\,d) reveal a lower performance for longer lead
times; this is probably due to minimum 3-hourly instead of 1-hourly lagged
response values employed as covariates in the persistency model. The circular
distributional tree and forest seem to partially compensate for the lower skill of
the lagged response values by other covariates, hence their predictive
performance only slightly decreases for the longer lead time. This compensation
is especially evident for Innsbruck, where the performance difference between
the persistency model and the tree-based methods significantly increases from
the 1-hourly to the 3-hourly forecast.

\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  %\hspace{-2cm}
  %\includegraphics{boxplot_crpsraw.pdf}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores for wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$ and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    per for each of the $12$~months averaged over the $5$~validation years which yields a total of
    $288$~yearly mean values. The scores are shown for the climatological, 
    persistency and the linear model as well as for the circular distributional tree and
    forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}
